|

### 2021-03-10 paper/v2_taylor_francis/broadband.speed.covid.Rmd
Answered most points from reviewer 1 and point 7 from reviewer 2.

During the pandemic, working from home using digital technologies, whether partially
or exclusively, was transformed from a niche means of accessing work, albeit one that
had been on a slow, upward trend, to a widespread way of life in many countries.
The ability to work from home or telecommute meant millions retained their jobs and,
to a varying extent, maintained productivity during periods of strict
lockdowns around the world.
However, this ability has not been evenly distributed socially or spatially,
creating new intersections of economic and digital divisions.
On one side are those who can work from home, supported by digital technologies,
and have thus been able to enjoy both economic resilience and greater personal safety.
On the other side, previously employed individuals have been forced to accept
furlough or redundancy packages unless they are part of the cadre of essential
workers, who are potentially at high risk of infection.
Whilst the basis for this pandemic-generated divide has been viewed as mainly
occupational, here we consider whether it is also technological and, consequently, geographical.

<!-- our approach -->
Using the UK as a case study, this paper aims to understand how the quality and
reliability of internet service, as reflected in *experienced* internet speeds
during the spring 2020 lockdown,
may reinforce or redress the spatial and social dimensions of digital divisions.
We employ volunteered geographic data on individual broadband speed tests
and state-of-the-art time-series clustering methods to create clusters of UK local
authorities with similar temporal signatures of experienced internet speeds.
We then associate these clusters of local authorities with their socioeconomic
and geographic characteristics to explore how they overlap with or diverge from
the existing economic and digital geography of the UK.
Our analysis enables us to better understand how the spatial and social distribution
of education, occupation and online accessibility intersect to enable or hinder the
practice of telecommuting at a time of extreme demand.
We also consider what lessons can be learned from this time for a future where
telecommuting is likely to remain a more common means of accessing work,
at least in comparison to the pre-Covid era.

<!-- capability to WFH -->
The capability to work from home has previously been studied from the
perspective of whether work tasks in a given occupation both can be and are
allowed to be performed using digital technologies independently of
location or co-location with colleagues, including supervisors
[@allen2015effective; @singh2013modeling]. However, successful telecommuting also 
requires that the quality and reliability of digital services, particularly home 
internet connection speeds, enable the completion of work tasks with a minimum of 
delay or interruption. Prior to the pandemic, the performance of broadband services 
with respect to telecommuters was never tested at scale, as working from home and 
connecting to colleagues and workplace resources via the internet was the purview of 
a small minority of workers.
Instead, leisure use in the evening, when video streaming services are at their peak,
has been used to benchmark broadband performance and service delivery by different
Internet Service Providers (ISPs) [@ofcom2017].

Yet the shift towards telecommuting during various stages of lockdown around the world
has been drastic and there are speculations that post-Covid, the tendency to work from
home will be much higher, raising questions around whether internet
services can accommodate the increased demand.
For example, $47$% of people in employment in the UK worked solely from home in April
$2020$, whilst the same figure only reached $5$% the year before [@ons2020;
@ons2020lm2019].
A back of the envelope calculation suggests that up to 40% of the labour force
could work from home on an ongoing basis [@batty2020editorial].
Similar figures have been reported for other countries [@felstead2020homeworking].
Approximately $37$% of the European workforce worked from home in April $2020$ with
countries like Finland reaching $60$% [@eurofound2020].
In the US, almost half of the working population worked from home during the same
period because of the pandemic [@brynjolfsson2020covid],
and a recent estimate indicated that $37$% of all jobs in the US can be permanently
performed entirely from home [@NBERw26948].
None of these changes could have happened in the absence of reliable information
and communication technology (ICT) infrastructure -- both in terms of software
and hardware.
But while software innovations are easily diffused across space and
society^[See for example the huge success of videoconferencing apps such as Zoom
[@marks2020zoom].], the same does not apply for ICT hardware infrastructure such as
internet broadband connectivity.

<!-- ICT infrastructure and digital divides-->
The literature describes first level digital divides in terms of the
availability and quality of internet connectivity, such as that manifest in
different geographies in the UK [@riddlesden2014broadband; @philip2017digital].
Second level digital divides consider the presence or lack of the necessary skills
to effectively utilise digital technologies and the internet
[@blank2014dimensions; @van2011internet].
The third level focuses on the heterogenous returns of internet usage among
different socioeconomic groups and, consequently, how digital technologies can
assist in bridging or further enhancing existing socioeconomic divides.
[@stern2009levels; @van2014digital; @van2015third].
The capability to telecommute is related to all three levels of digital divides,
but more importantly leads to differentiated outcomes regarding the economic
resilience of people and places to overcome a systemic shock such as the current
pandemic.

As the quality of internet infrastructure and services, as well as the concentration
of different occupations are spatially dependent and clustered in space,
our approach offers a framework for understanding the impact of and interactions
between digital divisions geographically and socioeconomically.
By asking how resilient broadband speeds, and particularly upload speeds, are as
experienced in different parts of the UK during a time of extreme demand,
we interrogate which places benefit from the greater economic resilience
digital technologies can offer.
The structure of this paper is as follows.
First we review the literature on telecommuting and digital divides to better
understand their structural and spatial development pre-pandemic,
and thus their importance to the economic resilience of different places.
We then describe our data and methodology.
Our results section first offers a classification of how internet services vary
across clusters of UK local authorities and then assesses whether these clusters
replicate or repudiate other socio-economic and geographic patterns of economic
resilience.
We conclude with a discussion of the insights we have gained from our new
perspective on digital divisions.

@@

## Digital divides and economic resilience {#sec:2.2}
<!-- WFH and digital divides-->
Our understanding of telecommuting as a product of enabled occupations can be
described as a manifestation of the third level digital divide, as those who are able
to use digital technologies to work from home benefit from a high rate of return on
their use of the internet in terms of autonomy, flexibility, and time saved from
commuting [@peters2004employees; @siha2006telecommuting; @singh2013modeling].
These returns have been even greater during the Covid-19 crisis,
when those with the capability to telecommute also have the ability to maintain
their employment whilst protecting their health.
However, the success of these arrangements has been dependent upon the first
level digital divide, which is associated with access to and quality of internet
connectivity.
@SALEMINK2017360 provides a systematic review of the pre-pandemic,
first level digital divide in infrastructure quality between urban and rural areas
in various advanced economies.
Rural areas, predictably, fare worse, yet as @blank2018local highlight, variation
in individual internet uptake and use is a product of more complex spatial and 
demographic characteristics than simple rurality or urbanisation.
Yet whether this variation in infrastructure quality affects the spatial footprint of
telecommuting has not previously been investigated, in part because telecommuting has
not previously been a cause of concentrated demand and pressure on internet services.

There are indications that those who purchase high speed connections consume
more data of all sorts and use their connections for a greater variety of purposes
[@hauge2011consumer].
There is also a correlation between access to internet services and a
reduction in household transport spend [@bris2017ict].
Independently of whether the additional internet use and reduced travel is
because of increased telecommuting, these studies suggest that better internet
services enable households to make savings and efficiencies,
an example of the first level digital divide reinforcing the third level.

@@

## Experienced Broadband Speeds {#sec:3.2}
<!-- introduction of our broadband speeds data -->
To assess the internet quality and reliability across local authorities in the UK,
we utilise unique data comprising individual internet speed tests from
Speedchecker Ltd^[[https://www.broadbandspeedchecker.co.uk/](https://www.broadbandspeedchecker.co.uk/)].
This is a private company that allows internet users to check their own broadband
upload and download speeds, and stores every speed-check with timestamp and
geolocation information.
These data have been used before to assess digital divides [@riddlesden2014broadband]
and the impact of local loop unbundling regulatory processes
[@nardotto2015unbundling], and we followed the former's approach to remove outliers.
By using this volunteered geographic data, we are able to assess the internet speed
*experienced* by users, which may differ from the maximum speeds *advertised* by
ISPs.
We are particularly interested in upload speeds and the frequency of speed tests
over the period from March to May $2020$, as government statements indicate this is
when UK workers were first told to work from home if at all possible [@GovUK2020].
Average upload speeds are slower than average download speeds, at $9.3$Mb/s
mean upload speed for the whole sample,
compared to $29.6$Mb/s for download speeds, but they are also less associated
with internet-based, high-demand, leisure activities such as video streaming.
Therefore, upload speeds are more relevant to work-related activities such as
uploading documents or two-way audio, video, and text-based communication systems.

@@

However, there were insufficient observations -- only $631$ speed
tests per LAD on average -- for each for each working hour of each working day in
each Local Authority District (LAD) to profile speeds at that level of detail.
Spatial aggregation was also necessary because we could not follow individuals or
households and connect data points.
Therefore, we aggregate the $241,088$ individual, geolocated and time-stamped
speed-checks during the $13$ weeks of March to May inclusive for weekdays in $2020$
by each hour of the day and day of the week.
As our research aims to identify the geography of internet service resilience for
work purposes, bank holidays and the hours between midnight and $6:00$ were excluded,
as well as weekend days.
The composite week time-series thus comprise $18$ hours multiplied by $5$ weekdays or
$90$ time points per series.
The time-series were calculated for each of the $382$ LADs in the UK, standardised,
and then a *k*-means partitioning around medoids clustering algorithm
was applied using DTW.
We initially run the algorithm for $k \in \mathbb{N} \bigcap [5,15]$
and used cluster validity indices (CVIs) to pick the optimal solution of $k = 13$.
Following @sardatime the majority vote for the following CVIs was used: Silhouette
(max), Score function (max), Calinski-Harabasz (max), Davies-Bouldin (min), Modified
Davies-Bouldin (DB*, min), Dunn (max), COP (min).

@@

First, we control for the number of speed tests run per cluster inhabitant
between $9:00$-$10:59$ as well as the share of fast Virgin Media internet
connections. Regarding the former, we expect people in LADs with more unreliable
connections to test their internet speeds more often, and the results validate our
priors. Meanwhile, fast Virgin Media cable connections have historically
only been available to $45$% of premises in the UK [@ofcom2016], where the more
lucrative and competitive market originally attracted the cable TV provider.
Those in clusters $2$, $3$, $9$ and $11$ benefit from a higher proportion of
Virgin connections, which is an indication that people in these clusters are
more likely to live in urban areas, with more choice of broadband services.
In other words, they are more likely to be on the right side of the first level,
infrastructure-based digital divide, as we expected from the analysis in Section
[4.1](#sec:4.1).

>>>>

During the pandemic, working from home using digital technologies, whether partially
or exclusively, was transformed from a niche means of accessing work, albeit one that
had been on a slow, upward trend, to a widespread way of life in many countries.
The ability to work from home or telecommute meant millions retained their jobs and,
to a varying extent, maintained productivity during periods of strict
lockdowns around the world.
However, this ability has not been evenly distributed socially or spatially,
creating new intersections of economic and digital divisions.
On one side are those who can work from home, supported by digital technologies,
and have thus been able to enjoy both economic resilience and greater personal safety.
On the other side, previously employed individuals have been forced to accept
furlough or redundancy packages unless they are part of the cadre of essential
workers, who are potentially at high risk of infection.
Whilst the basis for this pandemic-generated divide has been viewed as mainly
occupational, here we consider whether it is also technological and geographical.

<!-- capability to WFH -->
The capability to work from home has previously been studied from the
perspective of whether work tasks in a given occupation both can be and are
allowed to be performed using digital technologies independently of
location or co-location with colleagues, including supervisors
[@allen2015effective; @singh2013modeling]. However, successful telecommuting also 
requires that the quality and reliability of digital services, particularly home 
internet connection speeds, enable the completion of work tasks with a minimum of 
delay or interruption. High levels of demand are one of the main causes of reduced 
reliability and slower connection speeds, as network bandwidth becomes congested. 
Yet prior to the pandemic, the performance of broadband services with respect to 
telecommuters was never tested at scale, as working from home and connecting to 
colleagues and workplace resources via the internet was the purview of a small 
minority of workers. Instead, leisure use in the evening, when video streaming 
services are at their peak, has been used to benchmark broadband performance and 
service delivery by different Internet Service Providers (ISPs) [@ofcom2017].

The shift towards telecommuting during various stages of lockdown around the world
has been drastic and there are speculations that post-Covid, the tendency to work 
from home will be much higher, raising questions around whether internet
services can accommodate the increased demand.
For example, $47$% of people in employment in the UK worked solely from home in April
$2020$, whilst the same figure only reached $5$% the year before [@ons2020;
@ons2020lm2019].
Similar figures have been reported for other countries [@felstead2020homeworking].
Approximately $37$% of the European workforce worked from home in April $2020$ with
countries like Finland reaching $60$% [@eurofound2020].
In the US, almost half of the working population worked from home during the same
period because of the pandemic [@brynjolfsson2020covid],
and a recent estimate indicated that $37$% of all jobs in the US can be permanently
performed entirely from home [@NBERw26948].
None of these changes could have happened in the absence of reliable information
and communication technology (ICT) infrastructure -- both in terms of software
and hardware.
But while software innovations are easily diffused across space and
society^[See for example the huge success of videoconferencing apps such as Zoom
[@marks2020zoom].], the same does not apply for ICT hardware infrastructure such as
internet broadband connectivity.

<!-- our approach -->
Using the UK as a case study, this paper aims to understand how the quality and
reliability of internet service, as reflected in *experienced* internet speeds
during the spring 2020 lockdown,
may reinforce or redress the spatial and social dimensions of digital divisions.
We employ volunteered geographic data on individual broadband speed tests
and state-of-the-art time-series clustering methods to create clusters of UK local
authorities with similar temporal signatures of experienced internet speeds.
We then associate these clusters of local authorities with their socioeconomic
and geographic characteristics to explore how they overlap with or diverge from
the existing economic and digital geography of the UK.
As the quality of internet infrastructure and services, as well as the concentration
of different occupations are spatially dependent and clustered in space,
our approach offers a framework for understanding the impact of and interactions
between digital divisions geographically and socioeconomically.
Our analysis enables us to better understand how the spatial and social distribution
of education, occupation and online accessibility intersect to enable or hinder the
practice of telecommuting at a time of extreme demand.
We also consider what lessons can be learned from this time for a future where
telecommuting is likely to remain a more common means of accessing work,
at least in comparison to the pre-Covid era.

By asking how resilient broadband speeds, and particularly upload speeds, are as
experienced in different parts of the UK during a time of extreme demand,
we interrogate which places benefit from the greater economic resilience
digital technologies can offer.
The structure of this paper is as follows.
First we review the literature on telecommuting and digital divides to better
understand their structural and spatial development pre-pandemic,
and thus their importance to the economic resilience of different places.
We then describe our data and methodology.
Our results section first offers a classification of how internet services vary
across clusters of UK local authorities and then assesses whether these clusters
replicate or repudiate other socio-economic and geographic patterns of economic
resilience.
We conclude with a discussion of the insights we have gained from our new
perspective on digital divisions.

@@

## Digital divides and economic resilience {#sec:2.2}
<!-- WFH and digital divides-->
The literature describes first level digital divides in terms of the
availability and quality of internet connectivity, such as that manifest in
different geographies in the UK [@riddlesden2014broadband; @philip2017digital].
Second level digital divides consider the presence or lack of the necessary skills
to effectively utilise digital technologies and the internet
[@blank2014dimensions; @van2011internet].
The third level focuses on the heterogenous returns of internet usage among
different socioeconomic groups and, consequently, how digital technologies can
assist in bridging or further enhancing existing socioeconomic divides.
[@stern2009levels; @van2014digital; @van2015third].
The capability to telecommute is related to all three levels of digital divides,
but more importantly leads to differentiated outcomes regarding the economic
resilience of people and places to overcome a systemic shock such as the current
pandemic.

Our understanding of telecommuting as a product of enabled occupations can be
described as a manifestation of the third level digital divide, as those who are able
to use digital technologies to work from home benefit from a high rate of return on
their use of the internet in terms of autonomy, flexibility, and time saved from
commuting [@peters2004employees; @siha2006telecommuting; @singh2013modeling].
In some aspects, these returns may have increased during the Covid-19 crisis,
when those with the capability to telecommute also have the ability to maintain
their employment whilst protecting their health.
However, the success of these arrangements has been dependent upon the first
level digital divide, which is associated with access to and quality of internet
connectivity.
@SALEMINK2017360 provides a systematic review of the pre-pandemic,
first level digital divide in infrastructure quality between urban and rural areas
in various advanced economies. Rural areas, predictably, fare worse.
This finding broadly applies in the UK, where studies that also analysed broadband 
speed checks concluded that average speeds are lower in rural areas, something that
has not been improved by policy measures to increase competition [@riddlesden2014broadband; @nardotto2015unbundling].
@riddlesden2014broadband found that levels of deprivation did not correlate with 
first level digital divides.

However, as @blank2018local highlight, variation
in individual internet uptake and use is a product of more complex spatial and 
demographic characteristics than simple rurality or urbanisation.
Dense urban areas were shown to suffer more from slow down during peak hours, 
although these services were more likely to be improved by increased competition 
between providers, such as between new entrants and Virgin Media cable connections
[@riddlesden2014broadband; @nardotto2015unbundling]. The latter were historically 
available to only $45$% of premises in the UK [@ofcom2016], where the more
lucrative and competitive market originally attracted the cable TV provider. 
Whether the variation in infrastructure quality and reliability affects the 
spatial footprint of telecommuting has not previously been investigated, in part 
because telecommuting has not previously been a cause of concentrated demand and 
pressure on internet services. Other intersections between digital divides have been
subject to study.
There are indications that those who purchase high speed connections consume
more data of all sorts and use their connections for a greater variety of purposes
[@hauge2011consumer].
There is also a correlation between access to internet services and a
reduction in household transport spend [@bris2017ict].
These studies suggest that better internet services enable households to make 
savings and efficiencies, an example of the first level digital divide reinforcing 
the third level.

@@

## Experienced Broadband Speeds {#sec:3.2}
<!-- introduction of our broadband speeds data -->
To assess the internet quality and reliability across local authorities in the UK,
we utilise unique data comprising individual internet speed tests from
Speedchecker Ltd^[[https://www.broadbandspeedchecker.co.uk/](https://www.broadbandspeedchecker.co.uk/)].
This is a private company that allows internet users to check their own broadband
upload and download speeds, and stores every speed-check with timestamp and
geolocation information.
These data have been used before to assess digital divides [@riddlesden2014broadband]
and the impact of local loop unbundling regulatory processes
[@nardotto2015unbundling], and we followed the former's approach to remove outliers.
By using this volunteered geographic data, we are able to assess the internet speed
*experienced* by users, which may differ from the maximum speeds *advertised* by
ISPs.
Indeed, those who seek to test their broadband are most likely to do so because 
they are experiencing slower speeds than expected, although any skew towards slower
speeds is balanced by the likelihood that those who test their broadband are also 
more ‘tech-savvy’ and / or have purchased higher speed packages that are not 
delivering the promised level of service.

We are particularly interested in upload speeds and the frequency of speed tests
over the period from March to May $2020$, as government statements indicate this 
encompasses the period when UK workers were told to work from home if at all possible [@GovUK2020].
Schools and various retail, leisure and hospitality businesses were closed from late 
March, and restrictions were gradually eased from late May.
Average upload speeds are slower than average download speeds, at $9.3$Mb/s
mean upload speed for the whole sample,
compared to $29.6$Mb/s for download speeds, but they are also less associated
with internet-based, high-demand, leisure activities such as video streaming.
Therefore, upload speeds are more relevant to work-related activities such as
uploading documents or two-way audio, video, and text-based communication systems.

@@

However, there were insufficient observations -- only $631$ speed
tests per Local Authority District (LAD) on average -- 
for each for each working hour of each working day in each LAD to profile speeds at 
that level of detail.
Spatial aggregation was also necessary because we could not follow individuals or
households and connect data points.
Therefore, we aggregate the $241,088$ individual, geolocated and time-stamped
speed-checks during the $13$ weeks of March to May inclusive for weekdays in $2020$
by each hour of the day and day of the week.
As our research aims to identify the geography of internet service resilience for
work purposes, bank holidays and the hours between midnight and $6:00$ were excluded,
as well as weekend days.
The composite week time-series thus comprise $18$ hours multiplied by $5$ weekdays or
$90$ time points per series.
The time-series were calculated for each of the $382$ LADs in the UK, standardised,
and then a *k*-means partitioning around medoids clustering algorithm
was applied using DTW.
We initially run the algorithm for $k \in \mathbb{N} \bigcap [5,15]$
and used cluster validity indices (CVIs) to pick the optimal solution of $k = 13$.
Following @sardatime the majority vote for the following CVIs was used: Silhouette
(max), Score function (max), Calinski-Harabasz (max), Davies-Bouldin (min), Modified
Davies-Bouldin (DB*, min), Dunn (max), COP (min).

@@

First, we control for the number of speed tests run per cluster inhabitant
between $9:00$-$10:59$ as well as the share of fast Virgin Media internet
connections. Regarding the former, we expect people in LADs with more unreliable
connections to test their internet speeds more often, and the results validate our
priors. 
Those in clusters $2$, $3$, $9$ and $11$ benefit from a higher proportion of
Virgin connections, which is an indication that people in these clusters are
more likely to live in urban areas, with more choice of broadband services.
In other words, they are more likely to be on the right side of the first level,
infrastructure-based digital divide, as we expected from the analysis in Section
[4.1](#sec:4.1).
